Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-07-18 paper1_jiturra.Rmd
review

Class relations can also be comprehended as the structure of social ties between classes as a characteristic of the social structure, represented as networks formed by different social classes [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001]. This means that friendship and family ties tend to be homogeneous concerning social status and demographic characteristics, while distant ties bridge individuals to different social groups and contribute to network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. Besides, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951]. However, the similarity of attitudes in segregated networks results more from the social differentiation that structures class ties than from individual preferences [@feld_focused_1981].

There are two approaches in the study of class relations from a network perspective. On the one hand, network diversity is defined as the degree of connectedness to dissimilar occupations, which represents vertical access to resource diversity embedded in social networks [@lin_access_1986]. However, diversity is defined as the variability of different ties within the network without necessarily providing a reference position to describe network similarity. On the other hand, network segregation is the absence of cross-class connections and represents how similar network ties are in reference to the individual. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirically, studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper-class forming connections with diverse individuals, in contrast to the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are observed in the composition of social ties, where the upper and intermediate classes maintain increasingly diverse and prestigious social environments compared to the working class [@cepic_how_2020;@carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that since class interests widen the social distance between proprietors and manual workers, the intermediate position of supervisors and their higher contact frequency with manual workers makes the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

[^altmodels]: Supplementary analyses employing alternative income inequality measures show that the results are robust when using the Inter-decile ratio (D9/D1) and the Top 10/Bottom 50 ratio. Additionally, I classified countries into low, middle-low, middle, middle-high, and high-income inequality groups based on quintiles according to the Gini index. Hence, I used country-fixed effects regressions to control for the cross-country differences and observed and unobserved societal characteristics. The results are consistent with the multilevel estimations.

>>>>

Class relations can be understood as the structure of social ties between different classes within the social structure, represented by networks formed by various social classes [@blau_macrosociological_1977]. Empirical evidence consistently shows homophily in social relations within the social network literature [@mcpherson_birds_2001]. This indicates that friendships and family ties are often homogeneous regarding social status and demographic characteristics, whereas distant ties connect individuals to different social groups and enhance network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. Moreover, socialization preferences do play a role in the formation of segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, the similarity of attitudes within these networks arises more from the social differentiation structuring class ties than from individual preferences [@feld_focused_1981].

The study of class relations from a network perspective can be approached in two ways. First, network diversity is defined as the degree of connectedness to dissimilar occupations, representing vertical access to resource diversity within social networks [@lin_access_1986]. In this context, diversity refers to the variability of different ties within the network without necessarily providing a reference point to describe network similarity. Second, network segregation, defined as the absence of cross-class connections, indicates how similar network ties are in reference to the individual. This perspective is conceptually closer to homophily, as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirical studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper class forming connections with diverse individuals, contrasting with the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are found in the composition of social ties, where the upper and intermediate classes maintain more diverse and prestigious social environments compared to the working class [@cepic_how_2020; @carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that class interests widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their higher contact frequency with manual workers make the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class. This suggests that limited life chances and lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

[^altmodels]: Supplementary analyses employing alternative income inequality measures show that the results are robust when using the Inter-decile (D9/D1) and the Top 10/Bottom 50 ratios. Additionally, I classified countries into low, middle-low, middle, middle-high, and high-income inequality groups based on quintiles according to the Gini index. Hence, I used country-fixed effects regressions to control for the cross-country differences and observed and unobserved societal characteristics. The results are consistent with the multilevel estimations.

### 2024-07-18 paper1_jiturra_supp-material.Rmd
review

>>>>

## Influential case analysis

```{r}
set1 <-  RColorBrewer::brewer.pal(n = 8, name = "Set1")
options(ggplot2.discrete.fill = set1)
options(ggplot2.discrete.colour = set1)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text=ggplot2::element_text(size=15,  family="serif"))

# dfreg$dclass3spo_V <- car::recode(dfreg$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
# df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
# sjPlot::tab_xtab(df1$dclass3spo_V,df1$PARTLIV)
dfreg <- na.omit(dfreg)

## Model -  EGP3 - DIGCLASS - V to Working -----------------------------$ 
digclas1_VW_0<- lmer(egal~1 + homclass3_V_res + (1|country2),data=dfreg,weights = WEIGHT)
digclas1_VW_1<- lmer(egal~1 + homclass3_V_res+know_total +female+agenum+partner+edyears+Q03pcm+workst+(1|country2),data=dfreg,weights = WEIGHT)
# digclas1_VW_2<- update(digclas1_VW_1, . ~ . +know_total) 
digclas1_VW_3 <- update(digclas1_VW_1, . ~ . + dclass3res_V)
# digclas1_VW_5 <- update(digclas1_VW_3, . ~ . + edyears+ Q03pcm+workst)

digclas1_VW_7 <- update(digclas1_VW_3, . ~ . +dclass3res_V*homclass3_V_res)

# Create an empty list to store results
results <- list()
countries<- levels(as.factor(dfreg$country2))
subset(dfreg,country2!="AUT")

# Iterate over each country
for (i in countries){
  # Exclude the i-th country
  subset_data <-dfreg %>% dplyr::filter(country2 != i)
  # Refit the model without the i-th country
  model_i <- lmer(egal~homclass3_V_res+know_total+female+agenum+partner+edyears + Q03pcm + workst + (1 | country2) + dclass3res_V + homclass3_V_res:dclass3res_V, data = subset_data,weights = WEIGHT)

  # Store the results
  results[[i]] <- model_i

  }

tidyresults <- list()
for (i in countries) {
  tidyresults[[i]] <- broom::tidy(results[[i]],conf.int=T)
}

jk<- bind_rows(tidyresults, .id = "country");jk$country <-paste("Excluding",jk$country) 
```

```{r byoneone-exclusion,fig.dim=c(10,10),fig.cap='Interaction of network homogeneity and social class on redistributive preferences. Excluding countries one by one.'}
jk %>% filter(term %in% c("homclass3_V_res",
                          "dclass3res_VIntermediate class (III+IV)","dclass3res_VWorking Class (V+VI+VII)",
                          "homclass3_V_res:dclass3res_VIntermediate class (III+IV)",
                          "homclass3_V_res:dclass3res_VWorking Class (V+VI+VII)")) %>% 
  ggplot(mapping = aes(y=estimate,x=term,ymin = conf.low, ymax = conf.high)) + 
  geom_point() +
  geom_errorbar(width = 0.2) +
  labs(x=NULL,y="Estimates") +
  scale_x_discrete(limits=rev(c("homclass3_V_res",
                          "dclass3res_VIntermediate class (III+IV)","dclass3res_VWorking Class (V+VI+VII)",
                          "homclass3_V_res:dclass3res_VIntermediate class (III+IV)",
                          "homclass3_V_res:dclass3res_VWorking Class (V+VI+VII)")),
                   labels=c("Hom:Working","Hom:Intermediate","Working Class","Intermediate Class","Homogeneity"))+
  geom_hline(yintercept = 0,color="red")+
  coord_flip() +
  facet_wrap(~country)
```

Using dfbetas and Cook distance, there are influential cases. Subsequently I see that in some cases the value is altered, but these changes are not significant.

```{r fig.dim=c(5,5),fig.cap='Cook distance'}
library(influence.ME)
estex.m31 <- influence(digclas1_VW_7, "country2")
dfbetas(estex.m31) 
cutpoint<- 2/sqrt(31) # cut point

plot(estex.m31, which="cook",
     cutoff=4/31, sort=TRUE,
     xlab="Cooks Distance",
     ylab="Country")
# sigtest(estex.m31, test=-1.96)
```
